Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-04-09 WholePlotTrajectories.tex
relecture

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the disturbance regime
that naturally shapes and maintains the structure, composition and
functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. To anticipate
the fate of tropical forests in the current context it is then urgent to
understand the response of communities to disturbance and the underlying
processes. Disturbance impact forest communities in changing both
abiotic and biotic environments, through modifications in the fluxes of
light, heat and water \citep{Goulamoussene2017}. Although in the
tropical context the forest environment is maintained after disturbance,
there is a succession of changes in the structure and functioning of
communities constituting a post-disturbance secondary succession. For
now, this succession has been largely studied through trajectories of
forest structural parameters such as aboveground biomass, tree height or
stem density \citep{Piponiot2016, Rutishauser2016}. Models based on
observed trajectories assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. Given the variety
of species response to disturbance and the huge diversity of tropical
forests, manifold post-disturbance biodiversity trajectories might
emerge \citep{Lindenmayer2012, Garcia_florez2017}. It is then essential
to anticipate post-disturbance trajectories and highlight the link
between biodiversity and disturbance, specifically the link between
disturbance intensity and frequency and the rate, endpoint and duration
of the trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. The link between biodiversity
and disturbance might be complicated by the huge diversity of tropical
communities and a variety of ecological processes may underlie the IDH
(colonisation, facilitation, coexistence,
etc)\citep{Lindenmayer2012, Garcia_florez2017}. There is a need to
clarify the scope of the IDH in terms of the different facets of
community diversity that are taxonomic richness, species abundance
distribution and community functioning. Besides, there is a need to
delineate the scope of the IDH in time, and highlight the moment when
the conditions for the IDH emerge \citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the species fostered by the disturbance
\citep{Lavorel2002, Bellwood2006}. Then, diversity metrics encompassing
taxonomic richness and evenness should be considered to assess the
changes in community abundance distribution. Besides, a functional
approach that accounts for species biological attributes and ecological
strategy would directly link community diversity, composition and
redundancy to ecosystem functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits that represent species ecological
strategy and determine how they respond to changing conditions
\citep{Diaz2005}. Specifically in tropical forests, the functional
approach revealed the post-disturbance deterministic processes entailing
a shift from a dominance of ``conservative'' slow-growing species
dealing with scarce resources, to a dominance of ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{Rees2001, Reich2014, Herault2011}. This shift is translated into
the trajectories of average community value of key functional traits
related to resource acquisition (leaf and stem traits) and life-history
strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines
communities resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) we defined taxonomic and functional post-disturbance
trajectories and examined the underlying ecological process, (ii) we
discussed the scope of the IDH regarding the different facets of
community diversity and the time after disturbance, and (iii) we
analyzed community resilience and time to recovery.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. The mean annual
temperature is 26\textdegree C. Elevation ranges from 5 to 50 m. Across
all plots the topography mainly corresponds to hilltops or hillsides,
while bottomlands cover less than 1 \% of the area. Plots are shallow
ferralitic acrisols over a layer of transformed saprolite with low
permeability and lateral drainage. Soil conditions are homogeneous, to
the exception of the highest hilltops where the thick surface allows a
free vertical drainage \citep{Gourlet-Fleury2004}.

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Rejou2017}. The three treatments constituted a disturbance
intensity gradient with increasing of above-ground biomass (AGB) lost
and surface disturbed.

@@

The taxonomic and functional trajectories were analysed from the
reference pre-disturbance inventories of 1984. The taxonomic diversity
was reported through species richness and the Hill number translation of
the Simpson index \citep{Hill1973}. These metrics allowed assessing the
taxonomic richness as well as evenness, through the comparison between
these two metrics: results will be discussed directly in terms of
taxonomic richness and evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015}.
The functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} the Rao index of quadratic entropy. The
Rao index combines species abundance distribution, and the average
pairwise functional dissimilarity between species computed by the gower
distance.

The impacts of initial disturbance were tested with the Spearman rank
correlation between the extremes of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the first two axes
from a PCA analysis, which lowered the weight of correlations among
traits as the PCA axes were combinations of most decoupled traits. For
each species, the traits probability density (TPD) were computed from
the mapping of individuals through two-dimension kernel density
estimators. Second, for each community, the TPD weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TPD was counted in each cell. The average count
across cells minus 1 returned the Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities the taxonomic composition changed towards more
species with acquisitive functional strategies. This translated into a
switch from high mean WSG in the community to high mean SLA and
chlorophyll content (see appendix I). The functional composition of
disturbed communities corresponded to cyclic compositional changes with
an incomplete recovery of the initial composition (Fig.
\ref{fig:NMDSplans}).

The maximum dissimilarity with the initial state was positively
correlated with the disturbance intensity for both taxonomic and
functional composition (\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial state was reached for taxonomic
composition between 15 to 25 years, for most of the plots, and around 22
years for functional composition.

@@

Community CWM average value of all traits and seed mass proportions
followed unimodal trajectories, either stabilizing or returning towards
their initial values, to the exception of leaf chlorophyll content,
which continued to increase for some T3 and T2 plots 30 years after
disturbance.

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
decreased and then slightly increased, but remained significantly lower
than their initial value (Fig. \ref{fig:CWM}). On the other side, bark
thickness and specific leaf area (\emph{SLA}) both increased. Bark
thickness remained substantially high after 30 years, and \emph{SLA} had
almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other lightest seed mass
classes (Supp. Mat. - Fig. S2).

@@

For undisturbed plots, taxonomic richness and Simpson diversity remained
stable over the 30 years of monitoring. In disturbed communities, after
low disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated with the disturbance intensity
(\(\rho_{Spearman}^{Richness}=0.50\)).

In all disturbed plots the Simpson diversity first increased until a
maximum reached after around 20 years. This maximum was positively
correlated with the disturbance intensity
(\(\rho_{Spearman}^{Simpson}=0.77\)). The Simpson diversity then
stabilized except for two T3 plots (plots 8 and 12) for which it kept
increasing (Fig. \ref{fig:DivTaxo}).

@@

The plot 7 from treatment 1 displayed constantly outlying functional
richness and Rao diversity and was removed from the graphical
representation for better readability. In undisturbed plots both
functional richness and Rao diversity remained stable along the 30
years. In disturbed plots, both trajectories depended on the disturbance
intensity, with their maximum being positively correlated to \%AGB loss
\(\rho_{Spearman}^{Richness}=0.76\) and \(\rho_{Spearman}^{Rao}=0.60\).
Functional richness and Rao diversity displayed for low disturbance
intensity a low but long-lasting increase up to a maximum reached after
20-25 years. Following high intensity, they displayed a fast but short
increase followed after 10 years by a slow decrease towards the initial
values.

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient (Fig. \ref{fig:IDHplot}). The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively).

@@

All disturbed plots had lower functional redundancy than control plots
and followed similar hump-shaped trajectories (\ref{fig:RedFunRest}).
The maximum redundancy loss was positively correlated with the
disturbance intensity (\(\rho_{Spearman}=0.47\)) and the recovery had
not attained initial values for any disturbed communities after 30
years.

@@

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with the disturbance leading a displacement on
the NMDS axis 1. The taxonomic compositional changes following
disturbance were similar among communities, which may correspond to the
recruitment of a group of pioneers, as \emph{Cecropia} spp.,
\emph{Vismia} spp., common to all communities whenever their initial
differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories continued with
a recovery of the initial composition which, althought not fully
achieved after 30 years, suggested the resilience of the taxonomic
composition resilience and the maintenance of the initial differences
among communities \citep{Folke2006}. The return towards initial
taxonomic composition showed that species not belonging to the
pre-disturbance community were rarely recruited, probably because of the
common dispersal limitations among tropical tree species
\citep{Svenning2005}.

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is believed to be the same as this of the initial community
\citep{Herault2018}. The changes in functional composition following
disturbance then relied upon the recruitment of species or functional
types previously infrequent or absent. These species were probably
competitive pioneers becoming dominant as the most efficient to benefit
from the light, space and nutrients made available by the disturbance.
This recruitment of pioneers led similar functional composition
trajectories for all plots, whenever the disturbance intensity. The
composition first went towards more resource-acquisitive strategies,
which was translated by a displacement to the right along the first axis
in the functional plane \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter, the pioneers recruited primarily were progressively excluded
by long-lived, more resistant and shade-tolerant species. The community
functional composition then returned towards more resource-conservative
strategies, suggesting the recovery of the initial functional
composition. This recovery translated in the functional plane by a
displacement left along the first axis and upward along the second axis
(Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
the taxonomic and functional resilience of communities. Taxonomicand
functional trajectories however appeared decoupled: while taxonomic
trajectories maintained the initial differences among communities, the
functional trajectories where similar and convergent in the functional
space \citep{Fukami2005}.

@@

Comparing community diversity along time after disturbance revealed
constrasting trajectories according to the distrubance intensity. Below
a disturbance intensity threshold, in the case of treatments 1 and 2,
the community taxonomic richness increased. In parallell, after a first
increase, the taxonomic eveness stabilized or strated to decrease after
30 years. Below the intensity threshold Contrastingly, above an
intensity threshold, in the case of treatment 3, the taxonomic richness
did not exceeded the initial value and the taxonomic evenness
continiously increased for 30 years without sign of return towards the
initial value. Below and above a threshold, corresponding to a
disturbance intensity between treatments 2 and 3, community taxonomic
richness and evenness changed markedly. A the disturbance intensity
threshold, the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity impacted the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. On the one hand, below a 20\%-25\% AGB loss,
the trees surviving after disturbance remained numerous enough to
maintain the high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. On the other hand, above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
then decreased following intense disturbance, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

>>>>

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural undisturbed
forests \citep{Morales-Hidalgo2015}. These threats affect the
disturbance regime that naturally shapes and maintains the structure,
composition and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. To anticipate
the fate of tropical forests in this context, understanding the response
of communities to disturbance and deciphering the underlying ecological
processes is of utmost importance. Disturbance impacts forest
communities in changing both abiotic and biotic environments, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}. Although the forest cover is generally
maintained after disturbance, there is a succession of changes in the
structure and functioning of ecological communities that defines a
post-disturbance ecosystem trajectory. For now, these trajectories have
been largely studied through trajectories of forest structural
parameters such as aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Recent works have identified some
of the determinants of post-disturbance biomass trajectories, like
pre-disturbance forest structure and composition or post-disturbance
environmental parameters \citep{Herault2018}. Regarding tree community
diversity and composition however, post-disturbance trajectories have
not been as thoroughly understood \citep{Guitet2018, Molino2001}. Given
the variety of species response to disturbance and the huge diversity of
tropical forests, manifold post-disturbance biodiversity trajectories
might emerge \citep{Lindenmayer2012, Garcia_florez2017}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}.

Moreover, we argue that there is a need to delineate the scope of the
IDH in time, using ecosystem time-explicit trajectories, and highlight
the moment when the conditions for the IDH emerge
\citep{Sheil2003, Shea2004}. Indeed, the link between biodiversity, IDH
and post-disturbance disturbance trajectories might be complicated by
the huge diversity found in tropical tree communities and the variety of
ecological processes that may underlie the IDH (colonization,
facilitation, coexistence,
etc)\citep{Lindenmayer2012, Garcia_florez2017}.

Analysing community response to disturbance requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the pool of species fostered (or hampered) by
disturbance \citep{Lavorel2002, Bellwood2006}. Then, diversity metrics
encompassing both richness and evenness should also be considered to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches because they shed light on the species biological
attributes that directly link community diversity, composition and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

The functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines
communities resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) we draw taxonomic and functional post-disturbance
trajectories and examine the underlying ecological process, (ii) we
discuss the scope of the IDH regarding the different facets of community
diversity and the time after disturbance, and (iii) we analyze community
resilience and time to recovery.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges from
5 to 50 m. Across all plots the topography mainly corresponds to
hilltops or hillsides, while bottomlands cover less than 1 \% of the
area. Plots are shallow ferralitic acrisols over a layer of transformed
saprolite with low permeability and lateral drainage. Soil conditions
are homogeneous, to the exception of the highest hilltops where the
thick surface allows a free vertical drainage
\citep{Gourlet-Fleury2004}.

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Rejou2017}. The three treatments were then transformed into a
continuous disturbance intensity gradient with increasing of
above-ground biomass (AGB) lost.

@@

The taxonomic and functional trajectories were analysed from the
reference pre-disturbance inventories of 1984. The taxonomic diversity
was reported through species richness and the Hill number translation of
the Simpson index \citep{Hill1973}. These metrics allowed assessing the
taxonomic richness as well as evenness, through the comparison between
these two metrics: results will be discussed directly in terms of
taxonomic richness and evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015}.
The functional diversity was reported using the functional richness and
functional evenness, through the Rao index of quadratic entropy. The Rao
index combines species abundance distribution, and the average pairwise
functional dissimilarity between species computed by the Gower distance.

The impacts of the initial disturbance levels on the maximum gain or
loss in richness and evenness were tested with the Spearman rank
correlation tests. Richness and evenness trajectories were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

Finally, functional redundancy was measured as the overlap among species
in community functional space \citep{Carmona2016}. First, the
individuals of the trait database were mapped in the plane of the first
two axes from a PCA analysis, which lowered the weight of correlations
among traits as the PCA axes were combinations of most decoupled traits.
For each species, the traits probability density (TPD) were computed
from the mapping of individuals through two-dimension kernel density
estimators. Second, for each community, the TPD weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TPD was counted in each cell. The average count
across cells minus 1 returned the Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories in disturbed plots over time.

The functional composition trajectory in disturbed communities resembled
to cyclic compositional changes with an incomplete recovery to the
initial composition (Fig.\ref{fig:NMDSplans}). The maximum dissimilarity
with the initial state was positively correlated with the disturbance
intensity for both taxonomic and functional composition
(\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial state was reached for taxonomic
composition between 15 to 25 years and around 20 to 25 years for
functional composition.

@@

In control plots, Community Weighted Means (CWM) of functional traits
remain stable in time.\\
In disturbed plots, they mostly followed unimodal trajectories, either
stabilizing or returning towards their initial values, to the exception
of leaf chlorophyll content, which continued to increase 30 years after
disturbance for some highly disturbed plots. Maximum height at adult
stage (\emph{Hmax}), leaf toughness and wood specific gravity
(\emph{WSG}) decreased in time and then slightly increased, but remained
significantly lower than their initial value (Fig. \ref{fig:CWM}). Bark
thickness and specific leaf area (\emph{SLA}) both increased in time.
Bark thickness remained substantially high after 30 years, and
\emph{SLA} had almost recovered to its initial value. Whatever the
functional traits, the maximum difference to initial value was highly
correlated to the disturbance intensity. Positive correlations were
observed for Leaf thickness, chlorophyll content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other lightest seed mass
classes (Supp. Mat. - Fig. S2).

@@

Taxonomic richness and evenness remained stable in control plots over
the 30 years of monitoring. In disturbed communities, after low
disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera. After intense disturbance the
taxonomic richness followed a more complex trajectory, decreasing for
ten years after disturbance before recovering to pre-disturbance values.
The maximum richness loss or gain after disturbance was positively
correlated with the disturbance intensity
(\(\rho_{Spearman}^{Richness}=0.50\)).

In all disturbed plots the evenness first increased until a maximum
reached after around 20 years. This maximum was positively correlated
with the disturbance intensity (\(\rho_{Spearman}^{Simpson}=0.77\)). The
evenness then stabilized except for two intensively-disturbed plots
(number 8 and 12) for which it kept increasing \ref{fig:DivTaxo}).

@@

Functional richness and evenness remained stable in control plots over
the 30 years of monitoring. In disturbed communities, both trajectories
depended on the disturbance intensity, with their maximum values in time
being positively correlated to \%AGB loss
\(\rho_{Spearman}^{Richness}=0.76\) and \(\rho_{Spearman}^{Rao}=0.60\).
Functional richness and evenness displayed for low disturbance intensity
a low but long-lasting increase up to a maximum reached after 20-25
years. For high disturbance intensity, they generally displayed a fast
but short increase followed after 10 years by a slow decrease towards
the initial values.

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) the taxonomic and functional diversity showed various
shape depending on the diversity indices and on the time since
disturbance (Fig. \ref{fig:IDHplot}). Regarding taxonomic diversity, the
relationship between disturbance intensity and diversity was more
markedly hump-shaped for richness than for evenness and peaked at 20\%
of initial AGB loss. Regarding functional diversity, the relationship
was almost linear and similar between richness and evenness. Generally,
all relationships were stronger 20 or 30 years after disturbance than
observed for 10 years.

@@

Control plots displayed stable functional redundancy over time. All
disturbed plots had lower functional redundancy than control plots and
followed similar hump-shaped trajectories (\ref{fig:RedFunRest}). The
maximum redundancy loss was positively correlated with the disturbance
intensity (\(\rho_{Spearman}=0.47\)) and the recovery trajectory had not
attained initial values for any disturbed communities after 30 years.

@@

Regarding taxonomic composition, the different tree communities were
substantially different before disturbance, and this is visualized by
their distinct starting points on the NMDS axis 2. These initial
differences were maintained all along the 30 years following
disturbance, with the disturbance leading a displacement on the NMDS
axis 1 only. Changes were thus similar and may correspond to the
recruitment of a group of pioneers, e.g. \emph{Cecropia} spp.,
\emph{Vismia} spp., shared by all plots and whatever their initial
taxonomic differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories seems to start
a recovery to the initial composition which, although far from fully
achieved after 30 years, suggested that taxonomic composition is
resilient and that the initial differences will be maintained in time
\citep{Folke2006}. This would show that species not belonging to the
pre-disturbance community were rarely recruited in the long-term,
probably because of the common dispersal limitations among tropical tree
species \citep{Svenning2005}.

Regarding functional composition, initial communities had similar
starting points and this contrasts to taxonomic composition. This means
that, despite differences in species composition, the functional
signature of the tree communities were quite similar. Following
disturbance, functional trajectories were consistent in the NMDS plane
with displacement intensities linked to disturbance intensities and thus
relied upon the recruitment of a shared pool of functional types
previously infrequent or absent. This common pool is composed of pioneer
resource-acquisitive (low leaf toughness, wood specific gravity, maximum
height and high Specific Leaf area) strategies that translated by a
displacement to the right along the first NMDS axis
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the pioneers
recruited primarily were progressively excluded by long-lived, more
competitive and shade-tolerant species. The community functional
composition then quickly returned towards more resource-conservative
strategies, suggesting the recovery of the initial functional
composition. This recovery translated in the functional plane by a
displacement left along the first axis and upward along the second axis
(Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
the taxonomic and functional resilience of communities. Taxonomic and
functional trajectories however appeared decoupled: while taxonomic
trajectories maintained the initial differences among communities, the
functional trajectories where similar and convergent in the functional
space \citep{Fukami2005} exemplifying the simultaneous operation of
trait-based assembly rules and species-level priority effects that
shapes tree community assembly in this forest, making it both
deterministic in functional space and historically contingent in
taxonomic space.

@@

Community trajectories depended on the disturbance intensity, with a
threshold of 20-25\% of AGB loss for which the taxonomic richness was
maximized and the taxonomic evenness remained resilient. Below a
disturbance intensity threshold (treatments 1 and 2), the taxonomic
richness increased and the taxonomic eveness increased first before
stabilizing or starting to decrease. In those cases, the trees surviving
after disturbance remained numerous enough to maintain the high
taxonomic richness of the pre-disturbance community \citep{Bongers2009}.
In parallel, the recruitment of pioneers, infrequent or absent before
disturbance, increased the taxonomic richness all the more so that the
disturbance was intense \citep{Martin2015, Chaudhary2016}. As these
pioneers became more dominant they balanced the usual hyper-dominance of
tropical forests and increased the taxonomic evenness
\citep{Baraloto2012a}. Contrastingly, above the intensity threshold
(treatment 3), the taxonomic richness did not exceed the initial value
and the taxonomic evenness continuously increased without sign of
reversal. Above the intensity threshold, the taxonomic richness of
surviving trees was too low to be offset by the recruitment of pioneers.
In the Guiana shield indeed, the pool of true pioneers specifically
recruited after disturbance is restricted to a few common genera (e.g.
\emph{Cecropia} spp., \emph{Vismia} spp.) \citep{Guitet2018}. At the
same time, pioneers became persistently dominant and prevented a return
towards the initial evenness and the recovery of hyper-dominant
shade-tolerant species.

After disturbance of intermediate intensity, taxonomic evenness
trajectories were hump-shaped, as already observed in the Guiana shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. Trajectories peaked at an intermediate time after
disturbance, around 15-20 years. This time corresponded to a balance
between first established pioneers and the more competitve
late-successional species.
